Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-15 Manuscript/manuscript.tex
Updates

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. For example, possible accumulation of ions at the membrane interface determines
the interaction affinity with charged species. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the best idealization of the ion binding as the direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens 
, which are in agreement with the previous interpretations of the experimental spectroscopic data.
On the other hand, the new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

Direct measurements of ion-membrane interactions in real biological systems are difficult. Simplified lipid bilayers are often used instead as entry-level models to shed light on the role of ions in complex biological membranes~\cite{scherer87,seelig90,cevc90}. For this reason, interactions of biologically relevant cations, especially Na$^+$ and Ca$^{2+}$, with zwitterionic phosphocholine (PC) bilayers have been widely studied in experiments~\cite{akutsu81, altenbach84, seelig90, cevc90, tocanne90, binder02, pabst07, uhrikova08} and classical MD simulations~\cite{bockmann03, bockmann04, Berkowitz12, melcrova16, javanainen17}. The details of ion binding are, however, not fully consistent in the literature. Non-invasive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering, and infrared spectroscopies mainly suggest that Na$^+$ ions exhibit negligible binding to PC lipid bilayers. On the contrary, Ca$^{2+}$ is observed to specifically bind to a couple of PC molecules using their phosphate groups~\cite{hauser76, hauser78, herbette84, akutsu81, altenbach84, binder02, pabst07, uhrikova08}. On the other hand, in general atomistic resolution molecular dynamics (MD) simulation models predict a stronger binding for the cations~\cite{catte16}. For example, simulations report Na$^+$ accumulating in the lipid interface~\cite{bockmann03} or Ca$^{2+}$ binding up to 4 PC lipids simultaneously, including not only interactions with phosphates but also with carbonyl oxygens~\cite{bockmann04, melcrova16, javanainen17}.
%{\color{red} Also, several experiments have also been interpreted to support the predictions from earlier MD simulations~\cite{bockmann03, vacha09a}. \todo{remove? Why we need this?}}

@@

In this work, we improve the cation binding of a POPC bilayer in MD simulations via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. For this, we apply a simple mean-field electronic continuum correction (ECC)~\cite{leontyev11} which implies scaling the partial charges of the atoms to account for the missing electronic polarizability in MD simulations~\cite{leontyev11}. Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. The newly developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while improving the membrane binding affinities significantly to sodium and calcium cations.

@@

According to ECC, the electronic polarizability can be modeled in non-polarizable MD by effectively embedding the ions in a homogeneous dielectric continuum with a dielectric constant $\epsilon_{el}$, which is the electronic part of the dielectric constant of the medium~\cite{leontyev11}. Following Coulomb's law,  ECC can be directly incorporated by scaling the charges with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$, giving
\begin{equation}
Q^{ECC} = f_q \cdot Q
\end{equation}
for the ECC corrected charges in simulation. 
Given that the  high frequency dielectric constant of water is $\epsilon _{el} = 1.78$ (\textit{i.e.,} the square of the refraction index), the scaling factor for ions in water is roughly $f_q \approx 0.75$. This scaling factor successfully improves the accuracy of simulations of solvated ions \cite{kohagen14, kohagen16, martinek17}, when quantitatively compared with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.
It is important to note that the 
high frequency dielectric constant 
is around 2 for almost any biologically relevant environment \cite{leontyev11}.
The dielectric discontinuity in a lipid bilayer arises
from the orientational polarization of the molecules. 
Thus, the same correction for the electronic polarizability can be 
applied throughout the lipid bilayer interface.

@@

To reproduce the experimental ion binding affinities,
we scanned the optimal interval for the scaling factor, $f_q~\in~(0.75, 1.0)$.
In general, scaling down the partial charges reduces the ion binding affinity 
and the experimentally measurable head group order parameter response to the bound charge~\cite{akutsu81, altenbach84, scherer89}. We found an optimal value $f_q = 0.8$, which is only slightly higher than the theoretical one ($f_q=0.75$). Directly applying the 0.8 scaling to the partial charges of the head group, the glycerol backbone, and the carbonyls reduces the area per lipid to 60~\AA$^2$. This area is smaller than the original in Lipid14 model ($65.6 \pm 0.5$~\AA$^2$)~\cite{dickson14} and the experimental value (64.3~\AA$^2$) \cite{kucerka11}. The decrease of the area per lipid arises from a reduced hydration of the lipid head group region after scaling charges which effectively reduces head group polarity. 
We solve this difficulty by reducing the effective radii of the modified atoms by lowering the $\sigma$ parameters in the Lennard-Jones potential by a factor of $f_\sigma = 0.89$. The same solution was taken for the ECC-ions in aqueous solvent \cite{kohagen14, kohagen16, Pluharova2014}. After reducing the $\sigma$ parameters, the area per molecule is restored to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Comparing MD simulation to NMR experiments, we can validate the simulation ion binding affinity to lipid bilayers using the ''electrometer concept''~ \cite{seelig87, catte16}. This method relies on the experimental observation that the C-H bond order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group (Fig.~\ref{simVSexpNOions}) are proportional to the amount of charge bound per lipid~\cite{seelig87}. Comparing this response is an excellent way to validate force fields as both NMR and MD can accurately determine order parameters ~\cite{catte16, ollila16}. The order parameters for all C-H bonds in lipid molecules can be accurately measured using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and membrane normal. Angular brackets point to the average over all sampled configurations.

@@

The measured change of the order parameter depends both on the head group response to the bound charge and on the amount of the bound charge (\textit{i.e.,} $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively).  The empirical factor $m_i$ has to be adequately quantified before the electrometer concept can be used to analyze the binding affinities. This calibration has been done experimentally for a wide range of systems~\cite{seelig87, beschiasvili91}. To calibrate the response of the head group order parameter to the bound charge in simulations, we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide  (DHAB) mixed with a POPC bilayer~\cite{scherer89}. DHAB\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
  -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit charge in the hydrophilic end. Thus, it is expected to locate in the bilayer similarly to the phospholipids, and its molar ratio then gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
NMR experiments report the used salt concentrations in two different ways when measuring head group order parameters. Some use the salt concentrations in water before solvating the lipids~\cite{akutsu81} while others use atomic absorption spectroscopy and report the salt concentration in the supernatant after the solvation of lipids~\cite{altenbach84}. In this work, we use the latter definition. In particular, we measure the salt concentration in the aqueous bulk region using the farthest point from the lipid bilayer in the aqueous phase. Notice that this definition differs from the one used by Catte et al.~\cite{catte16} where the former definition is used. Although these two definitions vary when applied to Ca$^{2+}$, they do not significantly affect the conclusions of this work.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-lipid model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} and its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Ref. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennart-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
\todo{Joe, please confirm that this is correct.}
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4). The simulation parameters used
in this work are summarized in Table~S1.
The simulated trajectories and parameters files are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC either the Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987}  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration time and was omitted from the analysis. Sufficient lipid neighbor exchange occurred during the simulations. Further details for the unscaled simulations and the trajectories are in Ref.~\citenum{catte16}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{kucerka11}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with the experimental values from \cite{ferreira13}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

@@

First, we present results for bilayers in pure water.
The ECC-lipids and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-lipids model
is $\approx$1\AA{} smaller
%\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
%SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-lipids model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI,
while still being close to the experimentally reported values.
We can thus conclude that the ECC-lipids model reproduces the experimental dimensions of the POPC
lipid bilayer with accuracy comparable to that for nother state of the art lipid models~\cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

Similarly, the acyl chain order parameters of the ECC-lipids model, as well as the Lipid14 model~\cite{dickson14}, agree with the experimental values within the error bars, as presented in Fig.~\ref{simVSexpNOions}. Notably, the experimentally measured forking and small order parameter values of $C_2$ segment in {\it sn}-2 chain are well reproduced by both models. This feature has been suggested to indicate that the carbonyl of {\it sn}-2 chain is directed towards the water phase, in contrast to the carbonyl in {\it sn}-1 chain, which would orient more along the bilayer plane~\cite{seelig75,schindler75,gawrisch92}. This arrangement may be a relevant feature for the ion binding details, which it is not fully reproduced by other available lipid models~\cite{ollila16}.
%This is expected as the ECC model does not modify the already highly optimized parameters of this region in Lipid14

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89}.
  }
\end{figure}

Before studying the sodium and calcium ion binding affinities, we have to quantify the response of the headgroup order parameters to the amount of bound charge by using mixtures of monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89}. These mixtures have a well-defined amount of bound charge per PC, i.e., the molar fraction of cationic surfactants. This direct relation results from the ability of dihexadecyldimethylammonium to be inserted in the lipid bilayers as any lipid because of its two hydrophobic acyl chains. Furthermore, available experimental data for these systems can be used to validate the sensitivity of lipid headgroup order parameters to the amount of bound charge in simulations \cite{scherer89}.

>>>>

\begin{abstract}
  % insert abstract here
Binding affinities and stoichiometries of Na$^+$ and Ca$^{2+}$ ions to phospholipid
bilayers are of paramount significance in the properties and functionality of cellular
membranes. For example, possible accumulation of ions at the membrane interface determines
the interaction affinity with charged species. Current estimates of binding affinities and
stoichiometries of cations are, however, controversial due to limitations in the available experimental and
theoretical methods. Experimentally one can assess these parameters by titrating
several membrane properties as a function of the ion concentration. However, the
interpretation of the experiments relies on theoretical models, which give
the best idealization of the ion binding as the direct experiments are not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic
resolution, therefore offering all the necessary information to interpret
experimental data without the need to resort to simplified models. However, the accuracy
of the available lipid models when interacting with ions is not sufficient for
such interpretation. In this work, we improve the binding details of Na$^+$ and Ca$^{2+}$ ions to the
1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic
polarization as a mean field correction, known as the electronic continuum correction (ECC).
This is applied to the partial charges of a selected state of the art POPC lipid model for MD simulations.
Our improved ECC-POPC model reproduces not only the
experimentally measured structural parameters for the ion-free membrane, but also the
response of lipid headgroup to a bound positive charge, and the binding affinities of 
Na$^+$ and Ca$^{2+}$ ions. With our new model we also observe negligible binding of Na$^+$ ions to POPC
bilayer and stronger interactions of Ca$^{2+}$ primarily with phosphate oxygens,
in agreement with the previous interpretations of the experimental spectroscopic data.
On the other hand, the new model results in Ca$^{2+}$ ions forming complexes with one to three POPC molecules with almost
equal probabilities, suggesting more complex binding stoichiometry than simple models
previously used to interpret the NMR data.
The results of this work pave the way to quantitative MD simulations of complex
biochemical systems with realistic electrostatic interactions in the vicinity of cellular
membranes.
\end{abstract}

@@

Direct measurements of ion-membrane interactions in real biological systems are difficult. Simplified lipid bilayers are often used instead as entry-level models to shed light on the role of ions in complex biological membranes~\cite{scherer87,seelig90,cevc90}. For this reason, interactions of biologically relevant cations, especially Na$^+$ and Ca$^{2+}$, with zwitterionic phosphocholine (PC) bilayers have been widely studied in experiments~\cite{akutsu81, altenbach84, seelig90, cevc90, tocanne90, binder02, pabst07, uhrikova08} and classical MD simulations~\cite{bockmann03, bockmann04, Berkowitz12, melcrova16, javanainen17}. The details of ion binding are, however, not fully consistent in the literature. Non-invasive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering, and infrared spectroscopies mainly suggest that Na$^+$ ions exhibit negligible binding to PC lipid bilayers. On the contrary, Ca$^{2+}$ is observed to specifically bind to a couple of PC molecules using their phosphate groups~\cite{hauser76, hauser78, herbette84, akutsu81, altenbach84, binder02, pabst07, uhrikova08}. On the other hand, most atomistic resolution molecular dynamics (MD) simulation models predict a stronger binding for the cations~\cite{catte16}. For example, simulations report Na$^+$ accumulating in the lipid interface~\cite{bockmann03} or Ca$^{2+}$ binding up to 4 PC lipids simultaneously, including not only interactions with phosphates but also with carbonyl oxygens~\cite{bockmann04, melcrova16, javanainen17}.
%{\color{red} Also, several experiments have also been interpreted to support the predictions from earlier MD simulations~\cite{bockmann03, vacha09a}. \todo{remove? Why we need this?}}

@@

In this work, we improve the cation binding of a POPC bilayer in MD simulations via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. For this, we apply a simple mean-field electronic continuum correction (ECC)~\cite{leontyev11} which implies scaling the partial charges of the atoms to account for the missing electronic polarizability in MD simulations. Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. The newly developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while improving the membrane binding affinities significantly to sodium and calcium cations.

@@

According to ECC, the electronic polarizability can be modeled in non-polarizable MD by effectively embedding the ions in a homogeneous dielectric continuum with a dielectric constant $\epsilon_{el}$, which is the electronic part of the dielectric constant of the medium~\cite{leontyev11}. Following Coulomb's law,  ECC can be directly incorporated by scaling the charges with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$, giving
\begin{equation}
Q^{ECC} = f_q \cdot Q
\end{equation}
for the ECC corrected charges in simulation. 
Given that the  high frequency dielectric constant of water is $\epsilon _{el} = 1.78$ (\textit{i.e.,} the square of the refraction index), the scaling factor for ions in water is roughly $f_q \approx 0.75$. This scaling factor successfully improves the accuracy of simulations of solvated ions, when quantitatively compared with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014, martinek17}.
It is important to note that the 
high frequency dielectric constant 
is around 2 for almost any biologically relevant environment \cite{leontyev11}.
The dielectric discontinuity in a lipid bilayer arises
from the orientational polarization of the molecules. 
Thus, the same correction for the electronic polarizability can be 
applied throughout the lipid bilayer interface.

@@

To reproduce the experimental ion binding affinities,
we scanned the optimal interval for the scaling factor, $f_q~\in~(0.75, 1.0)$.
The ion binding affinity was evaluated against experiments
using the head group order parameters and the electrometer concept \cite{seelig87,catte16},
as discussed more detail in the next section.
In general, scaling down the partial charges reduced the ion binding affinity.
We found an optimal binding affinity with the scaling factor of $f_q = 0.8$,
which is only slightly higher than the theoretical one ($f_q=0.75$).
Directly applying
the 0.8 scaling to the partial charges of the head group, the glycerol backbone, and
the carbonyls reduced the area per lipid to 60~\AA$^2$. This area is smaller than in the
original Lipid14 model ($65.6 \pm 0.5$~\AA$^2$)~\cite{dickson14} and in experiments
(64.3~\AA$^2$) \cite{kucerka11}. The decrease of the area per lipid arises from a
reduced hydration of the lipid head group region after scaling charges which effectively
reduces head group polarity. We solve this difficulty by reducing the effective radii of
the modified atoms by lowering the $\sigma$ parameters in the Lennard-Jones potential by a
factor of $f_\sigma = 0.89$. The same solution was taken for the ECC-ions in aqueous
solvent \cite{kohagen14, kohagen16, Pluharova2014, martinek17}. After reducing the $\sigma$ parameters,
the area per molecule is restored to the experimental value (Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Comparing MD simulation to NMR experiments, we can validate the ion
binding affinity in lipid bilayer simulations using the ''electrometer concept''~ \cite{seelig87, catte16}.
This method relies on the experimental observation that the C-H bond order parameters of $\alpha$ and $\beta$ carbons in a PC lipid head group (Fig.~\ref{simVSexpNOions}) are proportional to the amount of charge bound per lipid~\cite{seelig87}.
%Comparing this response is an excellent way to validate force fields as both NMR and MD can accurately determine order parameters ~\cite{catte16, ollila16}.
The order parameters for all C-H bonds in lipid molecules can be accurately measured using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and membrane normal. Angular brackets point to the average over all sampled configurations.

@@

The measured change of the order parameter depends on the head group response to the bound charge and on the amount of the bound charge (\textit{i.e.,} $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively).  The empirical factor $m_i$ has to be adequately quantified before the electrometer concept can be used to analyze the binding affinities. This calibration has been done experimentally for a wide range of systems~\cite{seelig87, beschiasvili91}. To calibrate the response of the head group order parameter to the bound charge in simulations, we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide  (DHAB) mixed with a POPC bilayer~\cite{scherer89}. DHAB\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
  -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit charge in the hydrophilic end. Thus, it is expected to locate in the bilayer similarly to the phospholipids, and its molar ratio then gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
NMR experiments report the used salt concentrations in two different ways when measuring head group order parameters.
Some use the salt concentrations in water before solvating the lipids~\cite{akutsu81}, $C_w$,
while others use atomic absorption spectroscopy and report the salt concentration in
the supernatant after the solvation of lipids~\cite{altenbach84}, $C_b$. In this work,
%we use the latter definition. In particular,
we measure the latter concentration in the aqueous
bulk region using the farthest point from the lipid bilayer in the aqueous phase.
Notice that former concentration was used by Catte et al.~\cite{catte16}.
Although these two definitions vary when applied
to Ca$^{2+}$, they do not significantly affect the conclusions of this work.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-lipid model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} and its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennart-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
\todo{Joe, please confirm that this is correct.}
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4). The simulation parameters used
in this work are summarized in Table~S1. Combability with openMM simulation package \cite{openmm7} was also
tested in SI. The simulated trajectories and parameters files are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC either the Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987}  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration time and was omitted from the analysis. Sufficient lipid neighbor exchange occurred during the simulations.
The trajectories and simulation files for unscaled systems are available from
Refs.~\citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{kucerka11} at {\color{red}??K}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with experiments \cite{ferreira13} at {\color{red}??K}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
    {\color{red}Joe, please add the temperatures}
  } 
\end{figure}

@@

First, we present results for bilayers in pure water.
The ECC-lipids and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-lipids model
is $\approx$1\AA{} smaller
%\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
%SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-lipids model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI,
while still being close to the experimentally reported values.
We can thus conclude that the ECC-lipids model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

Similarly, the acyl chain order parameters of the ECC-lipids model, as well as the Lipid14 model~\cite{dickson14}, agree with the experimental values within the error bars, as presented in Fig.~\ref{simVSexpNOions}. Notably, the experimentally measured forking and small order parameter values of $C_2$ segment in {\it sn}-2 chain are well reproduced by both models. This feature has been suggested to indicate that the carbonyl of {\it sn}-2 chain is directed towards the water phase, in contrast to the carbonyl in {\it sn}-1 chain, which would orient more along the bilayer plane~\cite{seelig75,schindler75,gawrisch92}.
While this arrangement may be a relevant feature for the ion binding details, it is not fully reproduced by other available lipid models~\cite{ollila16}.
%This is expected as the ECC model does not modify the already highly optimized parameters of this region in Lipid14

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
\end{figure}

Before studying the sodium and calcium ion binding affinities, we quantify the response of the headgroup order parameters to the amount of bound charge by using mixtures of monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89}. These mixtures have a well-defined amount of bound charge per PC, i.e., the molar fraction of cationic surfactants. This direct relation results from the ability of dihexadecyldimethylammonium to be inserted in the lipid bilayers as any lipid because of its two hydrophobic acyl chains. Furthermore, available experimental data for these systems can be used to validate the sensitivity of lipid headgroup order parameters to the amount of bound charge in simulations \cite{scherer89}.
